From physical principles to relativistic classical Hamiltonian and Lagrangian particle mechanics

### 2015-10-19 2015CompleteDerivation/completeDerivation.tex
Added detrev justification and first few math definitions

We start by fixing a \emph{physical system} we want to study, that is something we can interact with and perform measurements on (e.g. a cat, a planet, water). We call \emph{environment} everything else. We set what particular aspect we want to study (e.g. the trajectory when falling, the motion around a star, its flow in a pipe). We call \emph{state} a particular configuration in time of the portion under study. Since the state does not, in general, exhaust the description of the system, a part remains \emph{unstated}, and as such we'll call it, for lack of a better word. With this in mind, we introduce the following assumption:

@@

Much of the focus later will be given to the state under deterministic and reversible evolution. However, the non-deterministic part plays a fundamental part as well. In fact: it is this part of the evolution that determines what states are available and their description. Suppose we study the motion of a small particle, its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. Suppose we study the same small particle as suspended in a fluid, as it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states have. The issue is that the set of states has to invariant under the non-deterministic evolution as well. That is: if we apply the non-deterministic evolution to any state, we still need to end up in one of the possible states. If the Brownian motion is strong enough, the state gets knocked out of the set of states determined by a simple position/momentum pair.

A similar more drastic effect: consider a book and its motion under gravitation and inertial forces, its state being the position and momentum of the center of mass. As we increase the temperature of the air around the book, its motion is unaffected until, at some point, the book burns. Clearly, the non-deterministic evolution has brought one of the states outside the set.

As we have seen, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is: given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the individual trajectories are not (those fall within the unstated part). In fact: we cannot assume trajectory and states are defined at all for the unstated part. Consider a muon and its decay products: it is clear that the three outgoing particles have a state and trajectory of their own, it is clear that the mass and energy they have came from the muon, yet we cannot assign some internal state or trajectory for those parts of mass and energy before the muon decayed. What happens in the muon stays in the muon, so to speak.

@@

\textbf{Rationale}. While the choice of the object of our study (i.e. the system and the aspect we are interested in) is arbitrary, this assumption is at the heart of being able to write physics laws, to define physical states and to physically study a system. If we are to study a physical system through reliable experimentation and observation, we need deterministic and reversible processes to manipulate its state. We can provide different arguments that point in the same direction.

First, to be able to identify the system, we must first be able to isolate it from the environment. Intuitively, we can distinguish between two chairs because we can move the first to another room and sit on it without somehow having to drag the second as well. More precisely, we can manipulate the state of the first system without affecting the second, and vice-versa. If we extend this idea, to be able to properly call something a system we need to show that it is isolated: if I affect its state, the state of the other systems is not affected and vice-versa. In other words: the system future and past states are only determined by its own state.

Second, the aim of physics is to write laws that can be used to make prediction that can be validated experimentally. If I drop an anvil from a tower, it will accelerate at $9.81 m/s^2$; if I want the anvil to reach the ground at $x m/s$ I have to drop it from $y m$. To the extent that we want to make predictions in time, we need to have a one-to-one correspondence between initial and final states

@@

This link between state definition and deterministic processes should not be too surprising as the state, in the context of thermodynamics and system theory, is often defined as \emph{the set of variables needed to determine the future evolution of the system}. One source of confusion, though, may come from statistical process: those are not deterministic! The confusion is that the object of our study stops at the distribution (the ensemble), the part that can be indeed calculated, measured and prepared. Another source of confusion is that what is considered state changes depending on the context: the position of a molecule is part of the state for a single molecule system but not for a volume of gas. And again because in the latter context is not part of the deterministic motion of the system.

So, if it seems we need to require deterministic and reversible motion, why we call this an assumption? Why not call it requirement? Because it's a \emph{simplifying} assumption: it can never be totally achieved. A system can be prepared or measured up to a certain level of precision. Perfect isolation of a system is impossible both practically (e.g. black-body radiation, gravity, ...) and conceptually (e.g. if the system is perfectly isolated, we can't prepare it, we can't measure it, we can't study it: it's not a physical system). If the system is always interacting with the environment, then how can we get deterministic and reversible evolution? We can if the environment and the internal dynamics of the system interact in such a way that they affect and are little affected by the aspect we are studying. For example: the trajectory of a falling cat does not greatly affect and is not greatly affected by its digestive system or external temperature.

>>>>

We start by fixing a \emph{physical system} we want to study, that is something we can interact with and perform measurements on (e.g. a cat, a planet, water). We call \emph{environment} everything else. We set what particular aspect we want to study (e.g. the trajectory when falling, the motion around a star, its flow in a pipe). We call \emph{state} a particular configuration in time of the portion under study. Since the state does not, in general, exhaust the description of the system, a part remains \emph{unstated}, and as such we'll call it, for lack of a better word. With this in mind, we introduce the following:

@@

Much of the focus later will be given to the state under deterministic and reversible evolution. However, the non-deterministic part plays a fundamental role as well. In fact: it is this aspect of the evolution that determines what states are available and their description. Suppose we study the motion of a small particle, its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. Suppose we study the same small particle as suspended in a fluid, as it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states have. The issue is that the set of states must be closed under the non-deterministic evolution as well. That is: if we apply the non-deterministic evolution to any state, we still need to end up in one of the possible states. If the Brownian motion is strong enough, the state gets knocked out of the set of states determined by a simple position/momentum pair.

A similar more drastic effect: consider a book and its motion under gravitation and inertial forces, its state being the position and momentum of the center of mass. As we increase the temperature of the air around the book, its motion remains unaffected until, at some point, the book burns. Clearly, the non-deterministic evolution has brought one of the states outside the set, to the point that the system is no longer recognizable.

As we have seen, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is: given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the individual trajectories are not (those fall within the unstated part). In fact: we cannot assume trajectory and states are defined at all for the unstated part. Consider a muon and its decay products: it is clear that the three outgoing particles have a state and trajectory of their own, it is clear that the resulting total mass and energy came from the muon. Yet, before the decay, we cannot ascribe an internal state and trajectory to those parts. The state of a muon is not some combination of the state of an electron and two neutrinos.

@@

\textbf{Rationale}. The claim is that deterministic and reversible evolution is necessary to be able to study a physical system. That is: the only part of a system that can be studied through reliable and reproducible experimentation or observation is the part that undergoes deterministic and reversible evolution. We can provide different arguments that point in the same direction.

First, to be able to identify the system, we must first be able to tell it apart from the environment. Intuitively, we can distinguish between two chairs because we can move the first to another room and sit on it without having touched the second. We can manipulate the state of the first system without affecting the second, and vice-versa. So, to tell a system apart from its environment, there has to to be a certain amount of isolation between the two: if I affect its state, the state of the other systems is not affected and vice-versa. This means that the system future and past states are only determined by its own state and the state undergoes deterministic and reversible evolution.

Second, the aim of physics is to write laws that can be used to make prediction that can be validated experimentally. If I drop an anvil from a tower, it will accelerate at $9.81 m/s^2$; if I want the anvil to reach the ground at $x m/s$ I have to drop it from $y m$. To the extent that we want to make predictions in time, we need to have a one-to-one correspondence between initial and final states.

@@

This link between state definition and deterministic processes should not be too surprising as the state, in the context of thermodynamics and system theory, is often defined as \emph{the set of variables needed to determine the future evolution of the system}. As we saw before, this applies also to statistical processes: the distribution (the ensemble) as a whole can be indeed calculated, measured and prepared, not each individual element.

So, if it seems we need to require deterministic and reversible motion, why we call this an assumption? First of all, because it's an idealization: it can never be completely achieved in practice. A system can be prepared or measured up to a certain level of precision. Perfect isolation of a system is impossible both practically (e.g. black-body radiation, gravity, ...) and conceptually (e.g. if the system is perfectly isolated, we cannot interact with it: how can it studied through experimentation?). It's a simplifying assumption that can only be taken if the environment and the internal dynamics of the system interact in such a way that they affect and are little affected by the aspect we are studying. As we saw before, for example, assuming that the states consists of the position and momentum of the center of mass means assuming that the Brownian motion of the body is negligible.

The second reason is that technically the assumption comes first. If we start looking for a law, we are implicitly assuming that there exists one to be found. The opposite assumption, there is no law to be found, can't really be experimentally proven exhaustively (i.e. you can only show that you have not found it among a certain class). Very often the hard part in science is finding for what system and in what conditions such assumption hold. Finding the law after that is a much simpler endeavour, and is something that is routinely done by students in their intro labs.

We are now ready to capture the elements of our discussion and their properties through mathematical definitions.

\begin{defn}\label{statedef}
Fix a physical system for study. There exists a set $\mathfrak{S}$, which we call \emph{state space}, consisting of all the physically distinguishable configurations $\mathfrak{s}$, which we call \emph{states}. There exists a bijective map $f_{\Delta t}:\mathfrak{S} \leftrightarrow \mathfrak{S}$, which we call \emph{time evolution operator}, that links initial and final states.
\end{defn}

\textbf{Justification}. The collection of all possible states from a set. The deterministic and reversible evolution maps one element in the set to one and only one element of the set.$\Box$

In practice, we use physical quantities to identify states. For example, the state for an ideal gas is given by pressure, volume and temperature. To clarify whether we are referring to a particular value of such quantity, or the quantity itself, we introduce the following terminology:
\begin{center}
    \begin{tabular}{ | p{2.5cm} | p{5.5cm} | }
    \hline
    State variable & A quantity (discrete or continuous) that must be specified to identify a state (e.g. position). \\ \hline
    Label & A particular value for a state variable (e.g. position = $5m$). \\ \hline
    Label range & A set of possible values for a state variable (e.g. position = $[4.5m, 5.5m]$). \\
    \hline
    \end{tabular}
\end{center}

\begin{defn}\label{label}
Let $\mathbbm{I}$ be a set. Let $f_\mathbbm{I} : \mathfrak{S} \rightarrow \mathbbm{I}$ be a function that maps each state to an element of the set. We call the pair $(\mathbbm{I}, f_\mathbbm{I})$ a \emph{state variable}. We call the pair $(i, f_\mathbbm{I}^{-1}(i))$ a \emph{label}, where $i \in \mathbbm{I}$ and $f_\mathbbm{I}^{-1}(i) = \{\mathfrak{s} \in \mathfrak{S} | f_\mathbbm{I}(\mathfrak{s}) = i\}$ is the reverse image of $i$. We call the pair $(I, f_\mathbbm{I}^{-1}(I))$ a \emph{label range}, where $I \subset \mathbbm{I}$.
\end{defn}

\textbf{Justification}. A state variable associates each state to one of its values. A label, one of the possible values, can also be thought as representing all states that are identified by its value. A label range, a set of possible values, can also be thought as representing all states that are identified by those values. Abusing the notation, $\mathbbm{I}$ will represent both the set of possible values and the function, and $i$ both the possible value and the set of states.$\Box$

labels/state variables, cardinality of state variable, state id function (=1 for a particular state). Can we do for continuous variables as well? Can we find divergence free evolution for det/rev and independent variables?
